# Objects {#ch:objects}

## Vectors {#sec:vectors}

A vector is a simple, one-dimensional list of data, like a single column
in Excel or in SPSS. Typically a single vector holds a single variable
of interest. The data in a vector can be of various classes: numeric,
character (strings of letters, always enclosed in double quotes), or
logical (i.e., boolean, `TRUE` or
`FALSE`, may be abbreviated to
`T` or `F`).

-   `c`: Atomic data are combined into a vector by means of the
    `c` (combine, concatenate) operator.

-   `seq` The sequence operator, also abbreviated as a colon
    `:`, creates subsequent values.
```{r 2.1.1}
x <- 1:5
print(x)
2*(x-1)
```

Computations are also done on whole vectors, as exemplified above.
In the last example, we see that the result of the computation is
*not* assigned to a new object. Hence the result is displayed ---
and then lost. This may still be useful however when you use
R  as a pocket calculator.

-   `rep` Finally, the repeat operator is very useful in creating repetitive
    sequences, e.g. for levels of an independent variable.
```{r 2.1.2}
x <- rep( 1:5, each=2 )
x
```

## Factors {#sec:factors}

Factors constitute a special class of variables. A factor is a variable
that holds categorical, character-like data. R  realizes
that variables of this class hold categorical data, and that the values
are category labels or *levels* rather than real characters or digits,
as illustrated in the examples below.
```{r 2.2.1}
x1 <- rep( 1:4, each=2 ) # create vector of numbers
print(x1) # numeric
summary(x1) # of numeric vector
x2 <- as.character(x1) # convert to char
print(x2) # character vector
x3 <- as.factor(x1) # convert to factor
print(x3) # factor
summary(x3) # compare against summary(x1) above 
```

## Complex objects {#sec:complex.objects}

Simple objects, like the ones introduced above, may be combined into
composite objects. For example, we may combine all pancake ingredients
into a complex object of class `list`.

In R we often use a particular complex object, a *data
frame*, to hold various data together. A data frame is a complex object
like an Excel worksheet or SPSS data sheet. The columns represent
variables, and the rows represent single observations --- these may be
"cases" or sampling units, or single measurements repeated for each
sampling unit, depending on the study 
^[For repeated measures analyses, R does not require a multivariate or "wide" layout, with repeated measures for each participant on a single row, as SPSS does. Instead R always uses a univariate or "long" layout, with each measurement on a single row of input. See the `reshape` command to convert between layouts, discussed in the next chapter, section \@ref(sec:split.merge.reshape).].

The easiest way to create a data object is to read it from a plain-text
(ASCII) file, using the command `read.table`.
(Windows users must remember to use double backslashes in the file
specification string). An optional `header=TRUE`
argument indicates whether the first line contains the names of the
variables; argument `sep` specifies the
character(s) that separate the variables in the input file. The
`file` argument can be a string specifying a
local file, or a `url` to a web-based file, or a
call of function `file.choose()` to select a file
interactively. Argument `na.strings` specifies
the character string(s) that indicate missing values in the input file.
```{text}
# in Windows system
myexp <- read.table(
  file="f:\\temp\\mydata.txt", header=T, sep="," )
```

```{r 2.3.1}
nlspkr <- read.table(
  file=url("http://www.hugoquene.nl/emlar/intra.bysubj.txt"),
  header=TRUE, na.strings=c("NA","MISSING") )
```

It is also possible to read so-called CSV files (comma-separated values) saved from Excel or SPSS (`read.csv`), and it is also possible to read Excel or SPSS data files directly using extension packages (`readxl::readxl`, `foreign::read.spss`, see Chapter \@ref(ch:packages)).

The basic R  and extension packages already have many
datasets pre-defined, for immediate use. 
To see a long(!) overview of these datasets, enter the command `data()`.
