# Basic operations {#ch:basicoperations}

## Basics {#sub:basics}

### `<-`
This is the assignment operator: the expression to its right is
    evaluated (if applicable) and then assigned to the object on the
    left of the operator. Hence the expression
    `a<-10` means that the object `a`, a single
    number, "gets" the value of 10, i.e. the value of 10 is assigned to `a`. 
    The symbol resembles
    an arrow in the direction of assignment. The assignment may also be
    in the other direction, with symbol `->` (and see note 
    ^[The equal sign `=` is also available for assignment. Using
    it is somewhat dangerous, however, because the equal sign does not
    specify the direction of assignment explicitly.]). 
    There should be no space between the two characters making up the arrow. 
    Use spaces or brackets to avoid ambiguities and errors: 
```{r 3.1.1}
x <- 10 # assignment of atomic value 10 to object x
x < -10 # is value of x less than -10 ?
```

### `#`
indicates a comment: everything following this symbol, on the same
    line of input, is ignored.

### `scan`

This command reads a simple vector from the keyboard. Make sure to
    assign the result to a new object! Read in the numbers 1 to 10, and
    assign them to a new object.

### `objects`

This command shows a list of all objects in memory (similar to the
contents of the `Praat` Objects window). With `objects(pattern="abc")` the list is filtered so that only the objects matching the pattern string `"abc"` are shown.

### `rm` 
Objects are removed *forever* with this command.

### `print`

Contents of an object can be inspected with this command, or by just
    entering the name of the object, as in some examples above.

### `summary`

This command offers a summary of an object. The result depends on
    the data class of the object, as illustrated in section REF
    above.

### Workspace:

R  holds its objects in memory. The whole workspace,
    containing all data objects, can be stored from the 
    Rstudio window (`Session > Save Workspace As...`).
    This allows you to save a
    session, and continue your work later (`Session > Load Workspace...`).

### `save`

(to write) and

### `read`

(to read) an object from/to memory to hard disk. By default,
    R  data objects have the extension `.Rda`.

### No `undo`

Remember that there is *no* `undo` command, nor such a menu
    option. Save your work regularly. If in doubt, work with scratch
    copies of your data sets.

## Subselection {#sec:subselection}

Subselection within an object is a very powerful tool in
R. The subselection operator `x[…]` selects only those data from object `x`
that match the expression within square brackets. This expression can be a
single index number, a sequence or list of numbers, or an evaluated
expression, as illustrated in the following example.

In the following example, variable `x` contains
30 numbers, but 3 of these are `NA`. Notice that
the output of `is.na` is the input of
`table`.
```{r 3.2.1.pre, echo=FALSE}
set.seed(17)
x <- rgamma(30,10,10)
x[ sample(length(x),size=3) ] <- NA
```

```{r 3.2.1}
# is.na() returns TRUE/FALSE for each element of ’x’. 
# table() summarizes categorical data 
table( is.na(x) ) 
ok <- !is.na( x ) # exclamation mark means NOT 
which( !ok ) # which index numbers are NOT ok? inspect! 
mean( x[ok] ) # select ok values, compute mean, display 
```

Subselection can also be achieved by using the function
`subset(data, subset, select)`. The first
argument is the input data (set), the second argument is the selector
condition, and the optional third argument indicates which columns of a
data frame should be kept in the output.
```{r 3.2.2}
require(hqmisc)
data(talkers)
subset( talkers, subset=( age<45 & region=="W" ) )
```

This command selects rows from data frame `talkers` from the package `hqmisc`  
(see \@ref(ch:packages)) corresponding to speakers who are under 45 years of age, and who are from the West region.

## Split, merge, reshape {#sec:split.merge.reshape}

There are useful functions available to split and merge data frames.
First we create two example data frames. The first data frame has a list
of English vowels, with a phonological feature for each vowel, and with the average frequency of the second formant^[A formant is a resonant frequency of the vocal tract, counted from low to high. The second formant or F2 is related to the front--back articulatory dimension.] of each vowel [@pete52] spoken by male speakers.  The second data frame has a partially overlapping list of vowels, with key words by John Wells ^[See <http://en.wikipedia.org/wiki/Lexical_set>].

```{r 3.3.1}
vowelsymb <- c( "i","I","e","E","ae", "A","V","o","U","u", "@" )
v1df <- data.frame( vowel=vowelsymb,
                    feat=factor( c(rep("front",5),
                                   rep("back",5),
                                   "central" ) ),
                    F2=c( 2290,1990,NA,1840,1720, 
                          1090,1190,NA,1020,870,
                          NA ) )
v2df <- data.frame( vowel=vowelsymb[1:10], 
                    word=c("fleece","kit","face","dress","trap", 
                           "lot","strut","goat","foot","goose") )
```

### `split`

This command divides the data in the first argument (column `F2` in
    data frame `v1df`) into the groups defined by the second argument.
    
```{r 3.3.2}
with( v1df, split(F2,feat) ) -> v1list 
```

This is particularly helpful in combination with
`lapply` to apply a function to these grouped
data, e.g. to compute the mean of F2 for each vowel category
separately:

```{r 3.3.3.}
with(v1df, lapply( split(F2,feat), mean, na.rm=T ) ) 
```

Also see `unsplit`.

### `merge`

This command merges two data frames, by common columns. Specify
    argument `all=TRUE` if you want to include non-matching rows in the
    output, with NA's in the appropriate columns.
    
```{r 3.3.4}
m1 <- merge(v1df, v2df)
```

The resulting merged data frame is also sorted on the common
columns, unless argument `sort=FALSE`.

### `reshape` 

If you need to perform a
    Repeated Measures (within-subjects) analysis of variance (RM-ANOVA)
    in SPSS, your data have to be in "wide" data layout, with all
    observations from one subject on a single data line.
    R on the other hand uses the "long" data layout, with
    one observation per line, and with all descriptors of that
    observation repeated for each line. There is a convenient command
    `reshape` to convert data between the wide
    layout (of SPSS RM-ANOVA) and the long layout (of R ).
    To illustrate, first we read a wide data set:
```{r 3.3.3.1}
widedata <- read.table( 
  file=url("http://www.hugoquene.nl/emlar/widedata.txt"),
  header=T)
```

The wide data show the subject id, between-subject group, and three
    within-subject observations, for 6 subjects (with leading row
    numbers):
```{r 3.3.3.2}
head(widedata)
```

These data are then reshaped to long layout with the following
    command:
```{r 3.3.3.3}
longdata <- reshape( widedata, direction="long",
                     varying=c("item1","item2","item3"),
                     timevar="item", times=c("1","2","3"),
                     v.names="resp", idvar="subject")
```

The observations from multiple columns `varying` are
    collected into a new single column `v.names`, using
    identifiers in column `idvar`. The
    information contained in the multiple column names of
    `varying` is stored in a new single column
    `timevar`, using the values `times`. 
Inspect the two data frames to verify this.
