# Exploratory data analyses {#ch-EDA}

R  is more graphically oriented than most other
statistical packages; it relies more on plots and figures for initial
exploratory data analysis. Numerical summaries are of course also
available.

## `hist`

This command produces a histogram. There is a useful optional
    argument `breaks` to specify the number of
    bins (bars), or a vector of breaks between bins.

## `plot`

The default version of this command produces a scattergram of two
    variables. If you enter just one variable, then the index numbers of
    the observations are used on the horizontal axis, and the values on
    the vertical axis. Useful arguments are
    `title`, and `xlab`
    and `ylab` for axis labels. In addition, you
    can use a third variable to vary the plot symbols.

## `rug`

This command produces tick marks at the actual data values, yielding
    the visual effect of a rug. This is useful in combination with a
    scattergram or histogram. Try it out, with the following
    commands ^[The semicolon separates multiple R commands on a single line of input.]:
```{r 4.3.1}
x<-rnorm(100) 
hist(x,col="lightblue"); rug(x,col="darkblue")
```

## `boxplot`

This yields a
    boxplot summary [@tukey77] of one variable. You can also specify the
    dependent and independent variable, with argument
    `dv~iv1`; this will produce multiple
    boxplots for the dependent variable, broken down by the independent
    variable(s). 
    Two useful arguments for this command are:\
    `notch=T` to give additional information
    about the distribution, and\
    `varwidth=T` to scale the width of the boxes
    to the numbers of observations.
```{r 4.4.1}
boxplot( x, col="lightblue", horizontal=TRUE, notch=TRUE, varwidth=TRUE )
rug(x, col="darkblue" )
```

## `qqnorm`

This produces a quantile--quantile (QQ) plot. This plots the
    observed quantiles against the expected quantiles *if* the argument
    variable would be distributed normally. If the variable is indeed
    distributed normally, then the data should fall on an approximately
    straight line. Deviations of this line indicate deviations from
    normality. You can also add the expected line with
    `qqline`.

## `summary`
This command produces a numerical summary of the argument variable.
    However it does not supply standard measures of variability. We
    often need

## `var`

to compute the variance of the argument variable. Related functions
    are `sd` to compute standard deviation,
    `cov` to compute the covariance between two
    variables, and `cor` to compute their
    correlation.

## `length`

returns the length of the argument variable, i.e. the number of
    observations in that vector. This is useful for checking the number
    of data, as a preliminary for further analyses.
```{r 4.8.1}
valid.n <- function(x) { 
  length(x)-length(which(is.na(x))) 
  }
```

In the last command above, we have programmed a new function
    `valid.n`, using standard functions provided
    with R.

## `unique`

returns a vector containing the (unsorted) unique values of the
    argument variable, without duplicates. This is also useful for
    checking your data.
```{r 4.9.1}
x <- c(3,3,3,1,1,2); unique(x); sort(unique(x))
```

## `table`

returns a frequency table, i.e. reports the number of observations
    for each (combination of) value(s) of one or more variables. This
    again is helpful for checking your data, to inspect the numbers of
    observations per cell (see also \@ref(sec:subselection)). 
    Tables can be one-dimensional (see
    example below) or multidimensional. It is also possible to make a
    frequency table of a frequency table! 
    
```{r 4.10.1.a}
x <- rep(1:3,each=7); table(x) # three cells, each 7 observations, N=21
table( table(x) ) # ok: all three cells have 7 observations, N=21
```

A useful trick: if all cells should have the
    same number of observations, then the table-of-table should contain
    a single number representing the number of *cells*.
```{r 4.10.1.b}
x[ sample(1:3*7, size=2) ] <- NA # replace 2 out of 21 obs by NA
table( table(x) ) # not ok: not all three cells have 7 observations, N<21
```

The `lattice` package (see Chapter \@ref(ch-packages)) contains additional functions for higher-order data visualization: 
`xyplot`,
`histogram`,
`densityplot`, 
`cloud`,
and more. For more information, enter the command
`help(lattice)`.

---

Now that we have obtained some insightful figures, we might want to
include these in our documents. The best procedure in Rstudio is to go to the Plots tab in the lower right pane, and select `Export > Save as image...` or `Export > Save as PDF...`. Figures in PDF format (extension
`pdf`) are easy to include in LaTeX documents, and are best for
publication. Other formats (e.g. `png`) are easy to include
in LaTeX, MS Word and web documents.

