# Mixed-effects modeling {#ch-mixedeffects}

Many language (acquisition) studies are based on samples of two random
factors: a sample of participants (subjects) and a sample of language
items (words, sentences, texts). The two random factors are *crossed*,
i.e., each item is presented to each participant --- often only once, so
that a subject does not respond to the same item repeatedly in multiple
conditions. The analysis methods shown above
(`aov`, `lm`,
`glm`) all fail to acknowledge this particular
structure in the random part of the design. They include a single random
factor (named `Residual`) that aggregates all
random effects.

A better method is to use *mixed-effects modeling*,
which may be done in R by using the `lmer` command. Key
advantages of this method are (a) it allows multiple random
factors, crossed and/or nested, (b) it does not require homogeneity of
variance, (c) it is robust against missing data. Hence mixed-effects
modeling is quickly gaining in popularity
[@QB04; @QB08; @baay08; @HMS18].

For mixed-effects modeling, you need to install two add-on packages to
R , named `lme4` and `languageR` [@BDB08]. For more
information on packages, see Chapter \@ref(ch-packages) below. 
After activation of these packages, we
can simply perform a mixed-effects analysis. First, we read in an
example dataset [@QB08] in long data layout:

```{r 7.1.1}
x24 <- read.table( file=url("http://www.hugoquene.nl/emlar/x24r2.txt"), header=T )
```

These fictitious responses were provided by 24 subjects, for 36 items,
in 3 conditions, with rotation of items over conditions. This rotation
may be inspected for a small subset of the data frame:
```{r 7.1.2}
with( subset(x24, subj<=3&item<=6), table(subj,item,cond) ) 
```

Next, we need to specify that `cond` is a
categorical factor, and not a continuous predictor. In addition, we
specify the levels of the factor, we specify its contrasts, and indicate
that the second level is the baseline or reference level.

```{r 7.1.3}
x24$cond <- as.factor(x24$cond)
contrasts(x24$cond) <- contr.treatment( c(".A",".B",".C"), base=2 )
```

After these preliminary steps we can estimate an appropriate mixed-effects
model in a single command. The estimated model is also stored as an
object, and a summary is displayed.

```{r 7.1.4}
summary( x24.m1 <- lmer(resp ~ 1+cond+(1|subj)+(1|item),
                        data=x24, REML=FALSE) )
```

The output correctly shows that there are two unrelated random effects,
plus unexplained residual variance. Each response is now modeled as a
unique combination of the intercept (mean of baseline condition B), item
effect, subject effect, condition effect, and residual. The average
response in the baseline condition B is $0.046$ units. Responses in
condition A are 0.170 units higher than baseline, and in condition C
they are $-0.237$ units higher than baseline, i.e. $0.237$ units lower.

For reasons not discussed here ^[See Frequently Asked Questions about R, Question
7.35, at <http://cran.r-project.org/doc/FAQ/R-FAQ.html>], the significance levels of the
fixed effects are not reported in the output of
`lmer`. There are several solutions to obtain
these significance levels.

* The most conservative option [@HMS18] is to use the critical $t$ value
associated with the random effect that has the fewest levels (here
`subj`), corrected for the number of fixed
coefficients including the intercept (here $3$). 
If a fixed effect is significant by this very
conservative criterion, then it will also be significant by any other
criterion that is less conservative and more liberal.
```{r 7.1.5}
qt( p=1-.05/2, df=24-3 ) # critical value t*, alpha=.05, two-sided
```

Comparison of the fixed effects with this critical $t^*$ = `r round(qt( p=1-.05/2, df=24-3),2)` shows that
both conditions A and C differ significantly from the baseline condition
B.

* A second option is to estimate 95% confidence intervals for all
coefficients in the resulting mixed-effects model. This can be
time-consuming, as the mixed-effects model will be re-fit many times on
slightly varying datasets.
```{r 7.1.8}
# warnings may occur
print( x24.m1.ci <- confint( x24.m1, method="boot", nsim=250 ))
```

As the interval for `cond.A` is entirely
positive, we may conclude with 95% confidence that condition A yields
higher scores than the baseline condition B, and mutatis mutandis that
condition C yields lower scores than condition B.
